---
title: "Phase 2"
author: "Feng"
date: "11/21/2017"
output: 
  html_document: 
    theme: united
    highlight: tango
    toc: true
    toc_float: true
    number_sections: true
---
# Setup

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, warning = FALSE, include = TRUE, tidy = TRUE, message = FALSE, cache = TRUE)
# setwd('F:/Project_group/Git_Project/FHIR/CEL_FILE')

# basic setup for markdown, path information (beautiful idea from Haipeng!) and relevant packages

path <- "data/"
path.fmerc <- "data_fmerc/"
path.tbi <- "data_tbi/"
path.xrt <- "data_xrt/"


file.data <- "data_matrix.rds"
file.fmerc <- "FMERC.txt"
file.tbi <- "TBI.txt"
file.xrt <- "XRT.txt"

csv_sep <- "\t"

library(readr)
library(Biobase)
library(dplyr)
library(GEOquery)
library(simpleaffy)
library(genefilter)
library(magrittr)
library(annotate)
library(preprocessCore)
library(hgu133plus2cdf)
# qc
library(RColorBrewer) # load colors for plots
library(affyPLM) # pseudo images 
library(arrayQualityMetrics) # one-click chip diagnosis
```

<br>

# Select type

Beautiful idea from Haipeng, i like this. Note that the pval and logfc are for bicolor heatmap only, which might not be important for now. (consider get rid of it?)

```{r select_type, include=FALSE}
data_type <- "tbi"

if(data_type == "fmerc"){
    cel_path <-  paste(path,path.fmerc,sep="")
    cel_files <- file.fmerc
    data_matrix <- paste(cel_path,file.data,sep="")
    type_1 <- 24
    type_2 <- 7
    name_1 <- 'High'
    name_2 <- 'Low'
    dis_label_1 <- 'FMERC_High'
    dis_label_2 <- 'FMERC_Low'
    pvalue <- 0.5323591 # for bicolor heatmap
    logfc <- 0.5
    node_label <- 'label'
}else if(data_type == "tbi"){
    cel_path <-  paste(path,path.tbi,sep="")
    cel_files <- file.tbi    
    data_matrix <- paste(cel_path,file.data,sep="")
    type_1 <- 20
    type_2 <- 18
    name_1 <- '0_10'
    name_2 <- '65_80'
    dis_label_1 <- 'TBI_0_10' 
    dis_label_2 <- 'TBI_65_80' 
    pvalue <- 0.05
    logfc <- 0.6
    node_label <- 'label'
}else if(data_type == "xrt"){
    cel_path <-  paste(path,path.xrt,sep="")
    cel_files <- file.xrt
    data_matrix <- paste(cel_path,file.data,sep="")
    type_1 <- 9
    type_2 <- 12
    name_1 <- 'Acute'
    name_2 <- 'unAcute'
    dis_label_1 <- 'XRT_Acute'
    dis_label_2 <- 'XRT_unAcute'
    pvalue <- 7.308559e-32
    logfc <- 0.6
    node_label <- 'label'
}else{
    cel_path <-  paste(path,path.fmerc,sep="")
    cel_files <- file.fmerc
    data_matrix <- paste(cel_path,file.data,sep="")
    type_1 <- 24
    type_2 <- 7
    name_1 <- 'High'
    name_2 <- 'Low'
    dis_label_1 <- 'FMERC_High'
    dis_label_2 <- 'FMERC_Low'
    pvalue <- 2.423544e-37
    logfc <- 0.6
    node_label <- 'label'
}


```
<br>

# Load & Check

Load all avialable celfiles and then check if all of them could be used (which is obviously not- -)

```{r load_qc}
# this section load, and control the quality of each chip
# the aim is to identify the extreme outliers and kick them out= =


# FMERC should be match with data files
ph <- read.csv(paste(cel_path,cel_files,sep=""),sep=csv_sep) 
# in case for naming problem (perhaps only in linux?), the following are needed - 
# cfs <- paste0(cel_path, list.celfiles(cel_path))
# cfs_rename <- sapply(cfs, function(char){gsub(" ", "_", char)}) %>% unname()`
# mapply(file.rename,  from = cfs, to = cfs_rename)
celfiles <- read.affy(covdesc=cel_files, path=cel_path)

#### quality check ####
qc_dir <- paste0(cel_path, "qc")
dir.create(qc_dir)

# Outlier detection
arrayQualityMetrics(expressionset = celfiles,
                    outdir = paste0(qc_dir, "/pre_report_", data_type),
                    force = TRUE, do.logtransform = TRUE)

# Perform probe-level metric calculations on the CEL files:
celfiles.qc <- fitPLM(celfiles)

# make the pesudo image for each chip
dir.create(paste(qc_dir, "qc_pseImage", sep = "/"))
for (i in 1:sum(type_1 + type_2)){
        png_name <- paste(qc_dir, "qc_pseImage", 
                          celfiles.qc@phenoData@data$sample[i], sep = "/")
        png(filename = paste0(png_name, '.png') )
        image(celfiles.qc, which=i, add.legend=TRUE)
        dev.off()
}

```

# Delete bad .cel files and modify default params

This section delete the confounding and replicated samples from the ph object and re-read the celfiles and gcrma normalization.


```{r modify_data}
# identify data ids to remove
if (data_type == "fmerc"){
        remove.ids <- c(4, 17, 20, 19)
}
if (data_type == "tbi"){
        remove.ids <- c(10, 6)
}
# input a numeric vector (the sample ids), enter the ids manually

# to reset default params (from ph object)
ph <- ph[-remove.ids,]
write_tsv(ph, paste0(cel_path,"/", data_type,"_mod.txt"))
cel_files <- paste0(data_type, "_mod.txt")
celfiles <- read.affy(covdesc=cel_files, path=cel_path)

type_1 <- table(ph$disease_state)[1] %>% unname()
type_2 <- table(ph$disease_state)[2] %>% unname()
```


# Gcrma normalization and quality recheck

In this section we use gcrma normalization to process the purified microarrays. Next another round of arrayqualitymetrics is used to see if quality plots are consistent with expectation.


```{r gcrma_recheck}

gcnorm_filter <- function(cf){
  cf.gcrma <- gcrma(cf)
  cf.filtered <- nsFilter(cf.gcrma)
  print(cf.filtered$filter.log)
  return(cf.filtered)
}

cf.gf <- gcnorm_filter(cf = celfiles)
data.matrix <- cf.gf$eset %>% exprs

# recheck the data quality
arrayQualityMetrics(expressionset = cf.gf$eset,
                    outdir = paste0(qc_dir, "/post_report_", data_type),
                    force = TRUE, do.logtransform = TRUE)
# just check the html file
# question - what is the problem with the line trend?

```



# dim reduction PCA and T-SNE


```{r dim_reducing}
library(Rtsne)

# ph <- read.csv(paste(cel_path,cel_files,sep=""),sep=csv_sep)
# data.matrix <- readRDS(data_matrix)

data.PC = prcomp(t(data.matrix),scale.=FALSE)
# summary(data.PC)

make.pca <- function(dm = data.martix){
  data.PC = prcomp(t(data.matrix),scale.=FALSE)
  data.PC = summary(data.PC)
  PCdf <- data.PC$x %>%  as.data.frame() 
  PCdf
}

PCdf <- make.pca()
# sPCdf <- summary(PCdf)
color=ifelse(ph$recur == ph$recur[1], "green", "red")
pairs(~PC1 + PC2 + PC3,data = PCdf, 
      main="PCA", col = color) 

llabels <- ph$recur


set.seed(3000)
make.tsne <- function(df = t(data.matrix)){
  data.tsne <- Rtsne(df, dims = 2, perplexity=6, verbose=TRUE, max_iter = 500, check_duplicates = FALSE)
  tsne_df <- as.data.frame(data.tsne$Y)
  tsne_df
}

tsne.plot <- function(lb = llabels, dt = tsne_df){
  plot(dt, t='n', main="tsne")
  text(dt, labels=lb, col=color, cex = 0.8)
}

tsne_df <- make.tsne()
tsne.plot()

```


<br>





# hierarchical clustering for PCA and T-SNE

```{r hclust}
library(dendextend)
fseq <- colnames(data.matrix)

mod.df <- function(raw_df, fs = fseq, phdict = ph){
  rownames(raw_df) <- phdict$sample_name[match(fs, phdict$file_name)]
  raw_df
}

hclust_pipe <- function(df, method){
  distance <- dist(df, method = 'maximum')
  dendo <- hclust(distance) %>% as.dendrogram()
  lb <- labels(dendo)
  col <- ifelse(ph$disease_state[match(lb, ph$sample_name)]==dis_label_2, 'red', 'green')
  dendo %>% set('labels', lb) %>% 
  set("labels_col", col) %>% 
  set("labels_cex", 1) %>% 
  plot(main = paste("hierarchical clustering by",method) )
}

PCdf <- mod.df(PCdf)
tsne_df <- mod.df(tsne_df)
hclust_pipe(PCdf,method = 'PCA')
hclust_pipe(tsne_df, method = 't-SNE')
```



# limma for differential expression analysis

identify what genes are differentially expressed in the two groups

```{r limma}

library(limma)
library(hgu133plus2.db)
library(annotate)

DE_pipe <- function(cff = cf.gf){
  # make matrix
  samples <- ph$disease_state %>% as.factor()
  design <- model.matrix(~0 + samples)
  colnames(design) <- c(dis_label_1, dis_label_2)
  fit <- lmFit(exprs(cff$eset), design)
  if (data_type == "tbi"){
          contrast.matrix <- makeContrasts(TBI_65_80-TBI_0_10, levels=design)
  }
  else if (data_type == "fmerc"){
          contrast.matrix <- makeContrasts(FMERC_High-FMERC_Low, levels=design)
  }
  else if (data_type == "xrt"){
          contrast.matrix <- makeContrasts(XRT_Acute-XRT_unAcute, levels=design)
  }
  # compare
  huvec_fits <- contrasts.fit(fit, contrast.matrix)
  huvec_ebFit <- eBayes(huvec_fits)
  tab <- topTable(huvec_ebFit, coef=1,lfc = 0.5, number = 1000)   
}

anno <- function(chip = hgu133plus2.db, t = tab){
  gs2 <- AnnotationDbi::select(x = chip, keys = rownames(tab), keytype = 'PROBEID',
                             columns = c('ENTREZID','GENENAME','SYMBOL')) %>% tbl_df()
  tab$PROBEID <- rownames(tab)
  tab <- left_join(tab, gs2) %>% tbl_df()
}

tab <- DE_pipe()
tab <- anno()


tab <- tab %>% dplyr::select(logFC, P.Value, ENTREZID, GENENAME, SYMBOL, PROBEID )
#save(tab, data.matrix, ph,file = 'whp_data1.RData')

knitr::kable(head(tab))
```

# save our data
save the important data into .RData file, use data_type specific name for discrimination.

```{r save_data}
# save the needed to .RData for future load and comparative analysis # 
modVarType <- function(v1, c1, dt) {
  newc <- paste(c1, dt, sep = "_")
  assign(newc, v1, envir = .GlobalEnv)
  #browser()
}

cvarsToSave <- c("data.matrix", "ph", "tab", "data_type" )

lapply(cvarsToSave, function(cvar){
        var <- get(cvar)
        #browser()
        modVarType(v1 = var, c1 = cvar, dt = data_type)
})

save(list = paste(cvarsToSave, data_type, sep = '_'), 
     file = paste(data_type, "backup.RData", sep = "_"))


```



<br>

# bicluster heatmap

```{r heat}
library(d3heatmap)
library(preprocessCore)
library(gplots)

order_matrix <- function(data.matrix,ph){
  df_dem <- data.frame(t(data.matrix))
  df_dem$label <- ph$recur
  df_dem <- df_dem[order(df_dem$label),]
  df_dem$label <- NULL
  order_matrix <- t(as.matrix(df_dem))
  rownames(order_matrix) <- rownames(data.matrix)
  order_matrix
}

normf <- function(m = omatrix, de = tab){
  tab0 <- de[ order(-tab$logFC), ] 
  tab0 <- tab0[tab0$P.Value < 0.05 & abs(tab0$logFC) > 0.6,]
  de_pids <- tab0$PROBEID
  dem <- m[de_pids,]
  
  # colnames_dem <- ph$sample_name[order(ph$recur)]
  colnames_dem <- ph$sample_name[match(colnames(dem), ph$file_name)]
  rownames_dem <- de$ENTREZID[match(rownames(dem), de$PROBEID)]
  # normalize in 2D; be careful, if tab0 is null, there will be Error thrown
  dem <- t(normalize.quantiles(t(dem)))
  dem <- normalize.quantiles(dem)
  dimnames(dem) <- list(rownames_dem, colnames_dem)
  dem
}

omatrix <- order_matrix(data.matrix,ph)
dem <- normf()

png(filename = paste0(data_type, "_heatmap.png"))
heatmap.2(dem, Colv = F ,col = bluered, trace = "none", dendrogram = "row") #, Colv = F)
dev.off()
# d3heatmap(dem, scale = "none", dendrogram = "none")
```

```{r clear_workspace}
rm(list = ls())
```


# compare gene sets

```{r Venn}
load("fmerc_backup.RData")
load("tbi_backup.RData")
load("xrt_backup.RData")

require(gplots) 

## 
fmerc <- tab_fmerc$SYMBOL
tbi <- tab_tbi$SYMBOL
xrt <- tab_xrt$SYMBOL

# see intersect
ft_inter <- intersect(fmerc, tbi)

# plot Venn (raw)
venn(list(GrpF=fmerc, GrpT=tbi))


# find related pathways
export2reactome <- function(genelist, csv_name){
        df <- data.frame(gl = genelist)
        write_csv(df, paste(csv_name, "_de.csv"))
}
mapply(export2reactome, genelist = list(fmerc, tbi, xrt), csv_name = c("fmerc", "tbi", "xrt"))
export2reactome(ft_inter, csv_name = "ft_inter")


# find enriched GO terms
gProfileR::gprofiler(fmerc, max_p_value = 0.05)



```



